---
title: "LMM_fisheries"
author: "Medy Mu"
date: '2022-10-13'
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1. Input Data; Basic Data Processing
```{r}
x<-read.csv("fullSetWithMask.csv",as.is=TRUE,
            na.strings=c("NA",".",""," "))
dim(x) ## Check the dimension (number of rows and columns) of the data matrix.
head(x)
```


```{r}
## Drop Column of Row Numbers:
x<-x[,-1]
## Rename a Few Variables:
colnames(x)[colnames(x)=="taxa_grouped_weight"]<-"taxa"
colnames(x)[colnames(x)=="agremiacion_fed"]<-"fed"
colnames(x)[colnames(x)=="coop_edad_10"]<-"age"
colnames(x)[colnames(x)=="rr"]<-"region"
x$uniqueid<-as.factor(x$uniqueid)
x$fed<-as.factor(x$fed)
x$taxa<-as.factor(x$taxa)
x$region<-as.factor(x$region)
x$yearFactor<-as.factor(x$year)
x$monthFactor<-as.factor(x$month)
```

## 2. Basic Summaries
```{r}
table(x$taxa,useNA="always")
table(x$region,useNA="always")
length(unique(x$uniqueid))
length(unique(x$fed))
table(x$year)
summary(x$age)
summary(x$age[x$uniqueid=="00000-BCSCAM"])
x$age2<-((10*x$age) + (x$year - 2016))
summary(x$age2)
table(x$age2<0)
tbl<-table(as.character(x$uniqueid)[x$age2<0]); tbl
table(x$age2[x$uniqueid %in% names(tbl)],
      as.character(x$uniqueid[x$uniqueid %in% names(tbl)]),
      useNA="always")
## Naive Age Fix (the years associated w/ these samples may be incorrect):
x$age[(x$uniqueid=="00549-TAB")&(x$age2<0)]<-(x$age[(x$uniqueid=="00549-TAB")&(x$age2<0)] + 0.5)
x$age[(x$uniqueid=="00685-BCS")&(x$age2<0)]<-(x$age[(x$uniqueid=="00685-BCS")&(x$age2<0)] + 0.2)
x$age[(x$uniqueid=="00747-YUC")&(x$age2<0)]<-(x$age[(x$uniqueid=="00747-YUC")&(x$age2<0)] + 0.6)
x$age[(x$uniqueid=="00748-YUC")&(x$age2<0)]<-(x$age[(x$uniqueid=="00748-YUC")&(x$age2<0)] + 0.2)
x$age2<-((10*x$age) + (x$year - 2016))
summary(x$age2)
```

## 2.2 Taxa by coop

```{r}
tbl<-table(x$uniqueid,x$taxa)
dim(tbl)
summary(coopByTaxa<-as.numeric(tbl))
table(coopByTaxa==0)
coopByTaxa<-coopByTaxa[coopByTaxa>0]
hist(coopByTaxa,nclass=50,las=1)
```

## 3. Client's LME model


Note, from the nlme::ACF documentation: “This method function calculates the empirical autocorrelation function for the within-group residuals from an lme fit. The autocorrelation values are calculated using pairs of residuals within the innermost group level. The autocorrelation function is useful for investigating serial correlation models for equally spaced data.”

### 3.1 Full data set

```{r}
keep<-(!is.na(x$logcpue))
lme.out<-nlme::lme(logcpue ~ age + yearFactor + monthFactor + region + taxa,
                   random = ~ 1|fed/uniqueid, data=x,subset=keep)
summary(lme.out)$tTable
acf.lme<-nlme::ACF(lme.out,maxLag=15)
plot(acf.lme,alpha=0.01/15)
```

```{r}
acf.lme<-acf.lme[-1,]
plot(acf.lme,alpha=0.005)
```

```{r}
qqnorm(fedn.re<-unlist(nlme::ranef(lme.out)$fed),las=1)
abline(a=0,b=sd(fedn.re),lwd=2,col=2)
```

```{r}
qqnorm(uid.re<-unlist(nlme::ranef(lme.out)$uniqueid))
abline(a=0,b=sd(uid.re),lwd=2,col=2)
```

### 3.2 Single taxa model

```{r}
keep<-((x$taxa=="OTRAS")&(!is.na(x$logcpue)))
otras.out<-nlme::lme(logcpue ~ age + yearFactor + monthFactor + region,
                     random = ~ 1|fed/uniqueid, data=x,subset=keep)
summary(otras.out)$tTable

acf.otras<-nlme::ACF(otras.out,maxLag=15)
plot(acf.otras,alpha=0.01/15)
```

```{r}
acf.otras<-acf.otras[-1,]
plot(acf.otras,alpha=0.005)
```

```{r}
qqnorm(fedn.re<-unlist(nlme::ranef(otras.out)$fed),las=1)
abline(a=0,b=sd(fedn.re),lwd=2,col=2)
```

```{r}
qqnorm(uid.re<-unlist(nlme::ranef(otras.out)$uniqueid))
abline(a=0,b=sd(uid.re),lwd=2,col=2)
```

## 4 Your mgcv Models
