---
title: "LMM_fisheries"
author: "Medy Mu"
date: '2022-10-13'
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(dplyr)
library(ggplot2)
library(lme4)
library(broom.mixed)
library(knitr)
library(patchwork)
library(nlme)
library(mgcv)
library(car)
```

## 1. Input Data; Basic Data Processing
```{r}
x<-read.csv("fullSetWithMask.csv",as.is=TRUE,
            na.strings=c("NA",".",""," "))
dim(x) ## Check the dimension (number of rows and columns) of the data matrix.
head(x)
```


```{r}
## Drop Column of Row Numbers:
x<-x[,-1]
## Rename a Few Variables:
colnames(x)[colnames(x)=="taxa_grouped_weight"]<-"taxa"
colnames(x)[colnames(x)=="agremiacion_fed"]<-"fed"
colnames(x)[colnames(x)=="coop_edad_10"]<-"age"
colnames(x)[colnames(x)=="rr"]<-"region"
x$uniqueid<-as.factor(x$uniqueid)
x$fed<-as.factor(x$fed)
x$taxa<-as.factor(x$taxa)
x$region<-as.factor(x$region)
x$yearFactor<-as.factor(x$year)
x$monthFactor<-as.factor(x$month)
```

## 2. Basic Summaries
```{r}
table(x$taxa,useNA="always")
table(x$region,useNA="always")
length(unique(x$uniqueid))
length(unique(x$fed))
table(x$year)
summary(x$age)
summary(x$age[x$uniqueid=="00000-BCSCAM"])
x$age2<-((10*x$age) + (x$year - 2016))
summary(x$age2)
table(x$age2<0)
tbl<-table(as.character(x$uniqueid)[x$age2<0]); tbl
table(x$age2[x$uniqueid %in% names(tbl)],
      as.character(x$uniqueid[x$uniqueid %in% names(tbl)]),
      useNA="always")
## Naive Age Fix (the years associated w/ these samples may be incorrect):
x$age[(x$uniqueid=="00549-TAB")&(x$age2<0)]<-(x$age[(x$uniqueid=="00549-TAB")&(x$age2<0)] + 0.5)
x$age[(x$uniqueid=="00685-BCS")&(x$age2<0)]<-(x$age[(x$uniqueid=="00685-BCS")&(x$age2<0)] + 0.2)
x$age[(x$uniqueid=="00747-YUC")&(x$age2<0)]<-(x$age[(x$uniqueid=="00747-YUC")&(x$age2<0)] + 0.6)
x$age[(x$uniqueid=="00748-YUC")&(x$age2<0)]<-(x$age[(x$uniqueid=="00748-YUC")&(x$age2<0)] + 0.2)
x$age2<-((10*x$age) + (x$year - 2016))
summary(x$age2)
```

## 2.2 Taxa by coop

```{r}
tbl<-table(x$uniqueid,x$taxa)
dim(tbl)
summary(coopByTaxa<-as.numeric(tbl))
table(coopByTaxa==0)
coopByTaxa<-coopByTaxa[coopByTaxa>0]
hist(coopByTaxa,nclass=50,las=1)
```

## 3. Client's LME model


Note, from the nlme::ACF documentation: “This method function calculates the empirical autocorrelation function for the within-group residuals from an lme fit. The autocorrelation values are calculated using pairs of residuals within the innermost group level. The autocorrelation function is useful for investigating serial correlation models for equally spaced data.”

### 3.1 Full data set

```{r}
keep<-(!is.na(x$logcpue))
lme.out<-nlme::lme(logcpue ~ age + yearFactor + monthFactor + region + taxa,
                   random = ~ 1|fed/uniqueid, data=x,subset=keep)
summary(lme.out)$tTable
acf.lme<-nlme::ACF(lme.out,maxLag=15)
plot(acf.lme,alpha=0.01/15)
```

```{r}
acf.lme<-acf.lme[-1,]
plot(acf.lme,alpha=0.005)
```

```{r}
qqnorm(fedn.re<-unlist(nlme::ranef(lme.out)$fed),las=1)
abline(a=0,b=sd(fedn.re),lwd=2,col=2)
```

```{r}
qqnorm(uid.re<-unlist(nlme::ranef(lme.out)$uniqueid))
abline(a=0,b=sd(uid.re),lwd=2,col=2)
```

### 3.2 Single taxa model

```{r}
keep<-((x$taxa=="OTRAS")&(!is.na(x$logcpue)))
otras.out<-nlme::lme(logcpue ~ age + yearFactor + monthFactor + region,
                     random = ~ 1|fed/uniqueid, data=x,subset=keep)
summary(otras.out)$tTable

acf.otras<-nlme::ACF(otras.out,maxLag=15)
plot(acf.otras,alpha=0.01/15)
```

```{r}
acf.otras<-acf.otras[-1,]
plot(acf.otras,alpha=0.005)
```

```{r}
qqnorm(fedn.re<-unlist(nlme::ranef(otras.out)$fed),las=1)
abline(a=0,b=sd(fedn.re),lwd=2,col=2)
```

```{r}
qqnorm(uid.re<-unlist(nlme::ranef(otras.out)$uniqueid))
abline(a=0,b=sd(uid.re),lwd=2,col=2)
```

## 4. Your mgcv Models

### EDA

```{r}
response_summary <- x %>%
    summarise(mean = mean(logcpue, na.rm=TRUE),
            median = median(logcpue, na.rm=TRUE),
            variance = var(logcpue, na.rm=TRUE),
            IQR = IQR(logcpue, na.rm=TRUE),
            sd = sd(logcpue, na.rm=TRUE),
            min = min(logcpue, na.rm=TRUE),
            max = max(logcpue, na.rm=TRUE))
knitr::kable(response_summary, caption = "Summary Statistics for \n the logcpue")
```

```{r}
ggplot(data = x, aes(x = logcpue)) +
  geom_histogram(fill = "darkgreen", color = "black") +
  labs(x = "logcpue", y = "Count",
       title = "Distribution of logcpue",
       caption = "Figure 1") +
  theme_bw() 
```

In this dataset, the response variable is `logcpue`, which represents catch per unit of effort, measured during different time points (Figure 1, Table 1). The distribution of this variable is unimodal and slightly skewed to the left. The center, defined by median, is 1.97. The spread, defined by the interquartile range (IQR), is 2.89. There are no evident outliers based on the histogram shown in Figure 1. To better capture the distribution of logcpue for individual cooperation, a random sample of 30 cooperation were selected, and their logcpue measured at different time points were plotted (Figure 2). Indeed, cooperation exhibit different patterns for the distribution of logcpue. Some shows a bimodel distribution, while some shows a unimodel distribution, This suggests that the distribution of logcpue differs based on cooperations, so we decided to include it as an individual level in our LMM model.

```{r, fig.width = 10, fig.height = 6}
set.seed(031622)
# get sample of 30 cooperations
  
sample_org <- x %>%
  distinct(uniqueid) %>%
  sample_n(30) %>% pull()
# get data for those cooperation
sample_data <- x %>%
  filter(uniqueid %in% sample_org)
# make a histogram of the response for each lemur
ggplot(data = sample_data, aes(x = logcpue)) +
  geom_histogram(fill = "darkgreen", color = "black") +
  facet_wrap(~ uniqueid, scales = "free") +
  labs(x = "logcpue",
       title = "Distribution of logcpue by randomly sampled 30 cooperation",
       caption = "Figure 2") +
  theme_bw() 
```

Other variables in the data that are of interests include year and month of each fishing event, region, taxa, and age of the federations. Figure 3 shows the relationships between year and logcpue for 30 randomly selected cooperations. Figure 4 shows the relationships between month and logcpue for 30 randomly selected cooperations. While some cooperations have a constant logcpue value across years or months, some cooperations have varying logcpue based on the year or month the fishery takes place, indicating that time alone cannot explain the distribution of logcpue observed and other factors during the year or month logcpue was collected may play a role in determining the pattern of logcpue. 

```{r, fig.width = 9, fig.height = 7}
ggplot(data = sample_data, aes(x = yearFactor, y = logcpue)) +
  geom_boxplot(fill = "darkgreen", color = "black") +
  facet_wrap(~ uniqueid, scales = "free") +
  labs(x = "Year",
       y = "logcpue",
       title = "logcpue vs year",
       caption = "Figure 3") +
  theme_bw()
```

```{r, fig.width = 9, fig.height = 7}
ggplot(data = sample_data, aes(x = monthFactor, y = logcpue)) +
  geom_boxplot(fill = "darkgreen", color = "black") +
  facet_wrap(~ uniqueid, scales = "free") +
  labs(x = "Month",
       y = "logcpue",
       title = "logcpue vs month",
       caption = "Figure 4") +
  theme_bw() 
```

Therefore, a new continuous variable that incorporated both year and month is created. 

```{r}
x$yearMonth <- ((x$year - min(x$year)) *12 + x$month)
```

In addition to year and month, region and taxa also play an important role in explaining logcpue. In particular, logcpue differs based on the region where the fisheries take place and on the taxa of the fish caught (Figure 5,6). Therefore, these two variables are worth including in the final models.


```{r}
region <- ggplot(data = x, aes(x = region, y = logcpue)) +
  geom_boxplot(fill = "darkgreen", color = "black") +
  labs(x = "Region", 
       y = "logcpue",
       title = "logcpue vs. region",
       caption = "Figure 5") +
  theme_bw() 
```

```{r}
taxa<- ggplot(data = x, aes(x = taxa, y = logcpue)) +
  geom_boxplot(fill = "darkgreen", color = "black") +
  labs(x = "Taxa",
       y = "logcpue",
       title = "logcpue vs. taxa",
       caption = "Figure 6") +
  theme_bw() 
```

```{r}
region + taxa
```

## Methods

Both models using mgcv and lme are included in this section.

Since catch per unit of efforts (cpue) were measured at different time points for each cooperation and that cooperations from the same federations tend to have similar distribution of logcpue, a multilevel model analysis was used to model the distribution of logcpue. The multilevel analysis was done at three levels. The first level included time points when cpue were collected; the second level included individual cooperations (``uniqueid``); and the third level included federations (``fed``). To test whether a multilevel model is necessary to capture the distribution of logcpue, an unconditional means model in which there are no predictors at any level was first fitted, and intraclass correlation was calculated to estimate the relative variability between cooperations and between federations (Table 2). Indeed, the intraclass correlation for cooperations is about `r round(1.093864^2/ (1.093864^2 + 1.189196^2 + 1.687456^2), 2)` and for federations is about `r round(1.189196^2/ (1.093864^2 + 1.189196^2 + 1.687456^2), 2)`, which means that the average correlation between any two responses from the same cooperations or from the same federations is about `r round(1.093864^2/ (1.093864^2 + 1.189196^2 + 1.687456^2), 2)` and `r round(1.189196^2/ (1.093864^2 + 1.189196^2 + 1.687456^2), 2)` respectively. This suggests that about 48% of the variability in the logcpue is explained by cooperation to cooperation variability and federation to federation variability. Knowing cooperation and federation, therefore, can explain almost half of the variability in the data, providing evidence that the multilevel model structure is useful in this setting.

```{r, cache=TRUE}
unconditional_means_mgcv <- bam(logcpue ~ 1 + s(fed, bs = "re") + s(uniqueid, bs = "re"), 
           data = x, na.action = na.omit)
summary(unconditional_means_mgcv)
```

```{r, cache=TRUE}
unconditional_means <- lme(logcpue ~ 1, 
                          list(fed = ~1, uniqueid = ~1), 
                           data = x, na.action = na.omit)
summary(unconditional_means)
```

Next, to understand the effect of time before adding other Level One covariates, the model that only includes year and month as Level One predictors was fitted (Table 3). The Pseudo $R^2$ estimating the change in within-cooperation variance between the unconditional means and the Model with Time is about 0.42. Therefore, understanding changes of time (i.e. years) accounts for 42% variability in logcpue.

```{r, cache=TRUE}
model_time <- lme(logcpue ~ yearMonth, list(fed = ~yearMonth, uniqueid = ~yearMonth), 
                  data = x, na.action = na.omit)
summary(model_time)
```

mgcv,however, cannot fit this model, which is one limitation of mgcv. We cannot encode a covariance term between two random effect in mgcv. It can only fit uncorrelated random effects. 

```{r}
# Pseudo R^2
sigm_moda <- 1.687456^2
sigm_madb <- 1.64283974 
r <- (sigm_moda - sigm_madb)/sigm_moda
r
```

Based on the EDA and model with only time as level one predictors, we decided to include `yearMonth`, `region`, and `taxa` as our level one predictors. `age2` was included as level three (federation) predictor since we hypothesized that the age of federation would impact the efficiency of catching fish (logcpue) of its subsidiary cooperation. Therefore, two models were fitted. One that includes interactions between age of federations and region and between age of federations and taxa, and one that does not include any interactions. AIC and BIC of both models were compared. Since both AIC and BIC are lower for model with additional interaction effects, providing evidence that model with interactions is a better model. One thing to note is that no random slopes were included in level two or three in both model because adding them make the model too complex and too computational heavy.

```{r}
model_no_int_mgcv <- bam(logcpue ~ yearMonth + region + taxa + age2 
                         + s(fed, bs = "re") + s(uniqueid, bs = "re"), 
           data = x, na.action = na.omit)
summary(model_no_int_mgcv)
```

```{r, cache=TRUE}
model_no_int <- lme(logcpue ~ yearMonth + region + taxa + age2, 
                     list(fed = ~1, uniqueid = ~1), 
                            data = x, na.action = na.omit)
summary(model_no_int)
```

```{r}
model_int_mgcv <- bam(logcpue ~ yearMonth + region + taxa + age2 
                         + age2:region + age2:taxa +
                         + s(fed, bs = "re") + s(uniqueid, bs = "re"), 
           data = x, na.action = na.omit)
summary(model_int_mgcv)
```

```{r, cache=TRUE}
model_int <- lme(logcpue ~ yearMonth + region + taxa + age2 + age2:region + age2:taxa, 
                     list(fed = ~1, uniqueid = ~1), 
                            data = x, na.action = na.omit)
summary(model_int)
```

Moreover, we have also tried adding different interactions in the model, such as interactions effects betwen yearMonth and region and between yearMonth and taxa as shown below. However, the model selection criterion by AIC and BIC is higher that than that for the previous model. Therefore, model with interactions between age2 and region and between age2 and taxa is our final model.

```{r}
model_int2_mgcv <- bam(logcpue ~ yearMonth + region + taxa + age2 
                         + yearMonth:region + yearMonth:taxa +
                         + s(fed, bs = "re") + s(uniqueid, bs = "re"), 
           data = x, na.action = na.omit)
summary(model_int2_mgcv)
```

```{r}
model_int2<- lme(logcpue ~ yearMonth + region + taxa + age2 + yearMonth:region + yearMonth:taxa, 
                     list(fed = ~1, uniqueid = ~1), 
                            data = x, na.action = na.omit)
summary(model_int2)
```


## Final Model

**Level 1 Model: Time**

$$ Logcpue_{ijk} = a_{ij} + b_{ij} yearMonth_{ijk} + c_{ij} region_{ijk} + d_{ij} taxa_{ijk} + \epsilon_{ijk}, \space \epsilon_{ijk} \sim N(0, \sigma^2)$$

**Level 2 Model: Cooperations**

$$a_{ij} = a_i + u_{ij}$$

$$b_{ij} = b_i $$

$$c_{ij} = c_i$$

$$d_{ij} = d_i$$

**Level 3 Model: Federations**

$$a_{i} = \alpha_0 + \alpha_1 age_{i} + \tilde u_{i}$$

$$b_{i} = \beta_0 $$

$$c_{i} = \gamma_0  + \gamma_1 age_{i}$$

$$d_{i} = \delta_0 + \delta_1 age_{i} $$


**Composite model**

$$ Logcpue_{ijk} = a_i + u_{ij} + b_i yearMonth_{ijk} + c_i region_{ijk} + d_i taxa_{ijk} + \epsilon_{ijk}, \space \epsilon_{ijk} \sim N(0, \sigma^2)$$


$$ = \alpha_0 + \alpha_1 age_{i} + \tilde u_{i} + u_{ij} + \beta_0 yearMonth_{ijk} + \gamma_0 region_{ijk} + \gamma_1 age_{i} region_{ijk} + \delta_0 taxa_{ijk}+ \delta_1 age_{i} taxa_{ijk} + \epsilon_{ijk}$$
$$\space \epsilon_{ijk} \sim N(0, \sigma^2)$$
$$ = \alpha_0 + \alpha_1 age_{i} + \beta_0 yearMonth_{ijk} + \gamma_0 region_{ijk} + \gamma_1 age_{i} region_{ijk} + \delta_0 taxa_{ijk} + \delta_1 age_{i} taxa_{ijk}$$
$$+ [\tilde u_{i} + u_{ij} + \epsilon_{ijk}] $$
$$\epsilon_{ijk} \sim N(0, \sigma^2),u_{ij} \sim N(0, \sigma_{u}^2), \tilde u_{i} \sim N(0, \sigma_{\tilde u}^2)$$

## Model Diagnostics

```{r}
anova(model_int_mgcv)
```

### Linearity 

```{r}
x_na <- na.omit(x)
```

```{r}
ggplot(data.frame(yearMonth=x_na$yearMonth,pearson=residuals(model_int_mgcv,type="pearson")),
      aes(x=yearMonth,y=pearson)) +
    geom_point() +
    theme_bw()
```

The residuals are randomly scattered, indicating that the data is linear and that there is a linear relationship between yearMonth and response.

## Normality

```{r}
qqnorm(residuals(model_int))
```

Points mostly fall along a straight diagonal line on the normal quantile plot, so we can safely assume that the data is normally distributed.

## Constant Variance

```{r}
plot(model_int)
```

The vertical spread of the residuals is not constant across the residual plot, suggesting that the constant variance condition is violated for linear mixed effects model. Therefore, some transformation should be done in the future to address this problem.

## Colinearity

```{r}
vif(model_int)
```

The VIF value for `age2` is over 5, which is potentially concerning. Future model thus should take colinearity into consideration in modeling logcpue.

## Prediction

```{r}
# generate prediction file
model <- model_int_mgcv
data <- x %>% 
  filter(is.na(logcpue)) 
################

df <- data %>%
  mutate(prediction = predict(model,data))%>% 
  select(c('prediction')) 
df <- cbind(index = rownames(df),df)
df$index <- as.numeric(as.character(df$index))
write.csv(df,'predictions.csv',row.names = FALSE) # format: index,prediction
```
